Supplementary materials for: Modelling palaeoecological time series using generalized additive models

### 2018-08-15 supplementary-materials.tex
revise the materials

\setlength{\droptitle}{-2em}
  \title{Supplementary materials for: Modelling palaeoecological time series
using generalized additive models}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Gavin L. Simpson}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{May 13, 2018}

@@

The following packages are required: \emph{mgcv}, \emph{scam},
\emph{ggplot2}, \emph{cowplot}, and \emph{tidyr}. Also, the
\emph{schoenberg} package is required; it is not on CRAN but can be
installed from Github.

@@

\begin{verbatim}
#> This is mgcv 1.8-23. For overview type 'help("mgcv-package")'.
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## schoenberg is not on CRAN, install from github:}
\NormalTok{## install.packages("devtools")}
\NormalTok{## devtools::install_github("gavinsimpson/schoenberg")}
\KeywordTok{library}\NormalTok{(}\StringTok{"schoenberg"}\NormalTok{)}
\KeywordTok{library}\NormalTok{(}\StringTok{"tidyr"}\NormalTok{)}

\NormalTok{## Default ggplot theme}
\KeywordTok{theme_set}\NormalTok{(}\KeywordTok{theme_bw}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

The example data sets are also stored on github;
\url{https://github.com/gavinsimpson/frontiers-palaeo-additive-modelling}.
Once downloaded the data are read in and processed a little

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## load braya so data set}
\NormalTok{braya <-}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"./data/braya-so/DAndrea.2011.Lake Braya So.txt"}\NormalTok{,}
                    \DataTypeTok{skip =} \DecValTok{84}\NormalTok{)}
\KeywordTok{names}\NormalTok{(braya) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Depth"}\NormalTok{, }\StringTok{"DepthUpper"}\NormalTok{, }\StringTok{"DepthLower"}\NormalTok{, }\StringTok{"Year"}\NormalTok{, }\StringTok{"YearYoung"}\NormalTok{,}
                  \StringTok{"YearOld"}\NormalTok{, }\StringTok{"UK37"}\NormalTok{)}
\NormalTok{braya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(braya, }\DataTypeTok{sampleInterval =} \NormalTok{YearYoung -}\StringTok{ }\NormalTok{YearOld)}
\KeywordTok{head}\NormalTok{(braya)}
\end{Highlighting}
\end{Shaded}

@@

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## fit small water GAM usin gamm() with a CAR(1)}
\NormalTok{mod <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small,}
            \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year), }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

The model summary is prepared from the \texttt{\$gam} component of the
fitted model

@@

The output shows the estimated complexity of the fitted smooth, in terms
of the effective degrees of freedom of the spline. An associated \(F\)
statistic and test of the null hypothesis of no trend (effect). Here the
estimated trend provides strong evidence against this null.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## plot CAR(1) process}
\NormalTok{maxS <-}\StringTok{ }\KeywordTok{with}\NormalTok{(small, }\KeywordTok{diff}\NormalTok{(}\KeywordTok{range}\NormalTok{(Year))) ## too large, truncate to 50}
\NormalTok{S <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{50}\NormalTok{, }\DataTypeTok{length =} \DecValTok{100}\NormalTok{)}

\NormalTok{car1 <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(}\KeywordTok{t}\NormalTok{(}\KeywordTok{outer}\NormalTok{(smallPhi, S, }\DataTypeTok{FUN =} \StringTok{`}\DataTypeTok{^}\StringTok{`}\NormalTok{)[}\DecValTok{1}\NormalTok{, , ])),}
                 \KeywordTok{c}\NormalTok{(}\StringTok{"Lower"}\NormalTok{,}\StringTok{"Correlation"}\NormalTok{,}\StringTok{"Upper"}\NormalTok{))}
\NormalTok{car1 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(car1, }\DataTypeTok{S =} \NormalTok{S)}

@@

The same model is fitted to the Braya-Sø data set. Note however that in
order to even fit the model with both a smooth and the CAR(1) process, I
have had to change the default optimiser used to fit the model, and
reduce the basis dimension to a small number. We also fit the model
using GCV, which is the defaut, hence no \texttt{method} argument

@@

The next couple of code chunks prepare plots of the fitted GAMS. The
general idea is to predict fom the fitted model for a fine grid of
points over the range of the time variable. First we plot the trend for
Small Water with an approximate 95\% confidence interval assuming
asymptotic normality

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{N <-}\StringTok{ }\DecValTok{300}   \CommentTok{# number of points at which to evaluate the splines}
\NormalTok{## Predict from the fitted model}
\NormalTok{newYear <-}\StringTok{ }\KeywordTok{with}\NormalTok{(small, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                             \DataTypeTok{length.out =} \DecValTok{200}\NormalTok{)))}
\NormalTok{newYear <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newYear,}
                 \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(mod$gam, newYear, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}
\NormalTok{newYear <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newYear,}
                     \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(}\DecValTok{2} \NormalTok{*}\StringTok{ }\NormalTok{se.fit),}
                     \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(}\DecValTok{2} \NormalTok{*}\StringTok{ }\NormalTok{se.fit))}

\NormalTok{## Plot simulated trends}
\NormalTok{small_fitted <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(newYear, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit)) +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year), }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{,}
                \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{fill =} \StringTok{"black"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data =} \NormalTok{small, }\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{d15N),}
               \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
\NormalTok{small_fitted}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## ggplot with data and fitted spline, then resids vs time in second panel}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                              \DataTypeTok{length.out =} \NormalTok{N)))}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newBraya,}
                  \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(braya.car1$gam, newBraya,}
                                     \DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}
\NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\DataTypeTok{df =} \KeywordTok{df.residual}\NormalTok{(braya.car1$gam))}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newBraya,}
                      \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit),}
                      \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit))}
\NormalTok{## add GAM GCV results}
\NormalTok{fit_gcv <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(braya.gcv, }\DataTypeTok{newdata =} \NormalTok{newBraya, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(newBraya, newBraya) }\CommentTok{# extend newBraya to take GCV results}
\NormalTok{newBraya[}\KeywordTok{seq}\NormalTok{(N}\DecValTok{+1}\NormalTok{, }\DataTypeTok{length.out =} \NormalTok{N, }\DataTypeTok{by =} \DecValTok{1}\NormalTok{), ]$fit <-}\StringTok{ }\NormalTok{fit_gcv$fit}
\NormalTok{newBraya[}\KeywordTok{seq}\NormalTok{(N}\DecValTok{+1}\NormalTok{, }\DataTypeTok{length.out =} \NormalTok{N, }\DataTypeTok{by =} \DecValTok{1}\NormalTok{), ]$upper <-}
\StringTok{    }\NormalTok{fit_gcv$fit +}\StringTok{ }\NormalTok{(}\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\KeywordTok{df.residual}\NormalTok{(braya.gcv)) *}\StringTok{ }\NormalTok{fit_gcv$se.fit)}
\NormalTok{newBraya[}\KeywordTok{seq}\NormalTok{(N}\DecValTok{+1}\NormalTok{, }\DataTypeTok{length.out =} \NormalTok{N, }\DataTypeTok{by =} \DecValTok{1}\NormalTok{), ]$lower <-}
\StringTok{    }\NormalTok{fit_gcv$fit -}\StringTok{ }\NormalTok{(}\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\KeywordTok{df.residual}\NormalTok{(braya.gcv)) *}\StringTok{ }\NormalTok{fit_gcv$se.fit)}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newBraya,}
                      \DataTypeTok{Method =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"GAMM (CAR(1))"}\NormalTok{, }\StringTok{"GAM (GCV)"}\NormalTok{), }
                                   \DataTypeTok{each =} \NormalTok{N))}

@@

\begin{center}\includegraphics[width=0.8\linewidth]{supplementary-materials_files/figure-latex/manuscript-fig-2-1} \end{center}

To proceed with the Braya-Sø example, we need to increase the basis
dimension (\texttt{k\ =\ 45}), fit using \texttt{method\ =\ "REML"}, and
use observational weights. Here I use the \texttt{sampleInterval}
variable as the measure of lake years per sample, and to avoid changing
the model likelihood, the weights are actually the values of
\texttt{sampleInterval} divided by the mean of \texttt{sampleInterval}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## TPRS, weights as sampleInterval}
\NormalTok{braya_reml <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{, }\DataTypeTok{bs =} \StringTok{"tp"}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
                  \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                  \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
\end{Highlighting}
\end{Shaded}

>>>>

\setlength{\droptitle}{-2em}

  \title{Supplementary materials for: Modelling palaeoecological time series
using generalized additive models}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Gavin L. Simpson}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{August 14, 2018}

@@

The following packages are required: \emph{mgcv}, \emph{scam},
\emph{ggplot2}, \emph{cowplot}, and \emph{tidyr}.

@@

\begin{verbatim}
#> This is mgcv 1.8-24. For overview type 'help("mgcv-package")'.
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"tidyr"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

In addition, the \emph{schoenberg} package is required; this
in-development package is not on CRAN but can be installed directly from
GitHub using functions from the \emph{devtools} package. To install the
package, install \emph{devtools} and then use
\texttt{devtools::install\_github()} to install \emph{schoenberg}, as
shown below:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## schoenberg is not on CRAN, install from github:}
\KeywordTok{install.packages}\NormalTok{(}\StringTok{"devtools"}\NormalTok{)}
\NormalTok{devtools::}\KeywordTok{install_github}\NormalTok{(}\StringTok{"gavinsimpson/schoenberg"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Assuming that the installation of \emph{schoenberg} completes without
error, the package can be loaded as usual

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(}\StringTok{"schoenberg"}\NormalTok{) }\CommentTok{# need to change the name of the package}
\end{Highlighting}
\end{Shaded}

The final environment-preparation step is to set the default
\emph{ggplot} theme, which the loading of \emph{cowplot} has
over-ridden. Here I use the more-minimal black-and-white theme
(\texttt{theme\_bw()})

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## Default ggplot theme}
\KeywordTok{theme_set}\NormalTok{(}\KeywordTok{theme_bw}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

\section{Load the data sets}\label{load-the-data-sets}

The example data sets are also stored on GitHub;
\url{https://github.com/gavinsimpson/frontiers-palaeo-additive-modelling}.
Once downloaded the data are read in and processed a little

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## load braya so data set}
\NormalTok{braya <-}\StringTok{ }\KeywordTok{read.table}\NormalTok{(}\StringTok{"./data/braya-so/DAndrea.2011.Lake Braya So.txt"}\NormalTok{,}
                    \DataTypeTok{skip =} \DecValTok{84}\NormalTok{)}
\NormalTok{## clean up variable names}
\KeywordTok{names}\NormalTok{(braya) <-}\StringTok{ }\KeywordTok{c}\NormalTok{(}\StringTok{"Depth"}\NormalTok{, }\StringTok{"DepthUpper"}\NormalTok{, }\StringTok{"DepthLower"}\NormalTok{, }\StringTok{"Year"}\NormalTok{, }\StringTok{"YearYoung"}\NormalTok{,}
                  \StringTok{"YearOld"}\NormalTok{, }\StringTok{"UK37"}\NormalTok{)}
\NormalTok{## add a variable for the amount of time per sediment sample}
\NormalTok{braya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(braya, }\DataTypeTok{sampleInterval =} \NormalTok{YearYoung -}\StringTok{ }\NormalTok{YearOld)}
\KeywordTok{head}\NormalTok{(braya)}
\end{Highlighting}
\end{Shaded}

@@

Instead of following the structure of the paper exactly, the subsequent
sections focus on the two example data sets in turn, beginning with
Small Water.

\subsection{Small Water}\label{small-water}

A GAM is typically fitted using the \texttt{gam()} function from the
\emph{mgcv} package. The typical call includes a formula describing the
response variable and the linear predictor separated by the
\texttt{\textasciitilde{}} (tilde) symbol. The linear predictor part of
the model contains the smooth function of the time variable in the data
set, and is indicated using the \texttt{s()} function. Unless other
options are needed, we only need specify where the variables in the
formula can be found via the \texttt{data} argument, and that we wish to
use REML smoothness selection.

Putting this together we have the following function call to fit a
simple GAM to the Small Water isotope data

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{m <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

As discussed in the paper, this model assumes that residuals\footnote{or
  equivalently that the observations, conditional upon the model, are
  independent.} are independent. To account for residual temporal
autocorrelation we can include a continuous-time first-order
autoregressive (CAR(1)) process in the model residuals for GAMs fitted
to data that are conditionally distributed Gaussian.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## fit small water GAM using gamm() with a CAR(1)}
\NormalTok{mod <-}\StringTok{ }\KeywordTok{gamm}\NormalTok{(d15N ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{small,}
            \DataTypeTok{correlation =} \KeywordTok{corCAR1}\NormalTok{(}\DataTypeTok{form =} \NormalTok{~}\StringTok{ }\NormalTok{Year), }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

The object returned by \texttt{gamm()} includes both the linear mixed
model and GAM faces of the model, and as a result we need to access the
separate elements (\texttt{lme} and \texttt{gam} respectively) when
proceeding to explore the model fit. The model summary is prepared from
the \texttt{\$gam} component of the fitted model

@@

The output shows the estimated complexity of the fitted smooth,
expressed in terms of the effective degrees of freedom of the spline. An
associated \(F\) statistic and test of the null hypothesis of no trend
(effect) are also shown. Here the estimated trend provides strong
evidence against this null.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{## plot CAR(1) process}
\NormalTok{S <-}\StringTok{ }\KeywordTok{seq}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{50}\NormalTok{, }\DataTypeTok{length =} \DecValTok{100}\NormalTok{)}
\NormalTok{car1 <-}\StringTok{ }\KeywordTok{setNames}\NormalTok{(}\KeywordTok{as.data.frame}\NormalTok{(}\KeywordTok{t}\NormalTok{(}\KeywordTok{outer}\NormalTok{(smallPhi, S, }\DataTypeTok{FUN =} \StringTok{`}\DataTypeTok{^}\StringTok{`}\NormalTok{)[}\DecValTok{1}\NormalTok{, , ])),}
                 \KeywordTok{c}\NormalTok{(}\StringTok{"Lower"}\NormalTok{,}\StringTok{"Correlation"}\NormalTok{,}\StringTok{"Upper"}\NormalTok{))}
\NormalTok{car1 <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(car1, }\DataTypeTok{S =} \NormalTok{S)}

@@

The next code chunk prepares a plot of the fitted GAMS. The general idea
is to predict from the fitted model for a fine grid of points over the
range of the time variable. Below I plot the trend for Small Water with
an approximate 95\% point-wise confidence interval that assumes
asymptotic normality

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{N <-}\StringTok{ }\DecValTok{300}   \CommentTok{# number of points at which to evaluate the smooth}

\NormalTok{## create new data to predict at; 200 evenly-spaced values over `Year`}
\NormalTok{newYear <-}\StringTok{ }\KeywordTok{with}\NormalTok{(small, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                             \DataTypeTok{length.out =} \DecValTok{200}\NormalTok{)))}
\NormalTok{## Predict from the fitted model; note we predict from the $gam part}
\NormalTok{newYear <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newYear,}
                 \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(mod$gam, newYear, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}
\NormalTok{## Create the confidence interval}
\NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\DataTypeTok{df =} \KeywordTok{df.residual}\NormalTok{(mod$gam))}
\NormalTok{newYear <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newYear,}
                     \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit),}
                     \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit))}

\NormalTok{## Plot estimated trend}
\NormalTok{small_fitted <-}\StringTok{ }\KeywordTok{ggplot}\NormalTok{(newYear, }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{fit)) +}
\StringTok{    }\KeywordTok{geom_ribbon}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{ymin =} \NormalTok{lower, }\DataTypeTok{ymax =} \NormalTok{upper, }\DataTypeTok{x =} \NormalTok{Year), }\DataTypeTok{alpha =} \FloatTok{0.2}\NormalTok{,}
                \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{fill =} \StringTok{"black"}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_point}\NormalTok{(}\DataTypeTok{data =} \NormalTok{small, }\DataTypeTok{mapping =} \KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \NormalTok{Year, }\DataTypeTok{y =} \NormalTok{d15N),}
               \DataTypeTok{inherit.aes =} \OtherTok{FALSE}\NormalTok{) +}
\StringTok{    }\KeywordTok{geom_line}\NormalTok{() +}
\StringTok{    }\KeywordTok{labs}\NormalTok{(}\DataTypeTok{y =} \NormalTok{d15n_label, }\DataTypeTok{x =} \StringTok{"Year CE"}\NormalTok{)}
\NormalTok{small_fitted}
\end{Highlighting}
\end{Shaded}

\subsection{Braya-Sø}\label{braya-s}

The same model is fitted to the Braya-Sø data set. Note however that in
order to even fit the model with both a smooth and the CAR(1) process, I
have had to change the default optimiser used to estimate the model
parameters, and reduce the basis dimension to a small number. We also
fit the model using GCV, which is the defaut, hence no \texttt{method}
argument

@@

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{N <-}\StringTok{ }\DecValTok{300}   \CommentTok{# number of points at which to evaluate the smooth}

\NormalTok{## data to predict at}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{with}\NormalTok{(braya, }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year =} \KeywordTok{seq}\NormalTok{(}\KeywordTok{min}\NormalTok{(Year), }\KeywordTok{max}\NormalTok{(Year),}
                                              \DataTypeTok{length.out =} \NormalTok{N)))}
\NormalTok{## add predictions from GAMM + CAR(1) model}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(newBraya,}
                  \KeywordTok{data.frame}\NormalTok{(}\KeywordTok{predict}\NormalTok{(braya.car1$gam, newBraya,}
                                     \DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)))}
\NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\DataTypeTok{df =} \KeywordTok{df.residual}\NormalTok{(braya.car1$gam))}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newBraya,}
                      \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit),}
                      \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit))}

\NormalTok{## add GAM GCV results}
\NormalTok{fit_gcv <-}\StringTok{ }\KeywordTok{predict}\NormalTok{(braya.gcv, }\DataTypeTok{newdata =} \NormalTok{newBraya, }\DataTypeTok{se.fit =} \OtherTok{TRUE}\NormalTok{)}
\NormalTok{crit.t <-}\StringTok{ }\KeywordTok{qt}\NormalTok{(}\FloatTok{0.975}\NormalTok{, }\KeywordTok{df.residual}\NormalTok{(braya.gcv))}
\NormalTok{newGCV <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{Year   =} \NormalTok{newBraya[[}\StringTok{"Year"}\NormalTok{]],}
                     \DataTypeTok{fit    =} \NormalTok{fit_gcv$fit,}
                     \DataTypeTok{se.fit =} \NormalTok{fit_gcv$se.fit)}
\NormalTok{newGCV <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newGCV,}
                    \DataTypeTok{upper =} \NormalTok{fit +}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit),}
                    \DataTypeTok{lower =} \NormalTok{fit -}\StringTok{ }\NormalTok{(crit.t *}\StringTok{ }\NormalTok{se.fit))}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{rbind}\NormalTok{(newBraya, newGCV)                }\CommentTok{# bind on GCV results}
\NormalTok{## Add indicator variable for model}
\NormalTok{newBraya <-}\StringTok{ }\KeywordTok{transform}\NormalTok{(newBraya,}
                      \DataTypeTok{Method =} \KeywordTok{rep}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\StringTok{"GAMM (CAR(1))"}\NormalTok{, }\StringTok{"GAM (GCV)"}\NormalTok{), }
                                   \DataTypeTok{each =} \NormalTok{N))}

@@

\subsubsection{Checking if the size of the basis expansion is
sufficient}\label{checking-if-the-size-of-the-basis-expansion-is-sufficient}

Clearly we are not able to identify both the wiggly trend and the CAR(1)
process in a single model. We might proceed by fitting a simple GAM via
\texttt{gam()} with REML smoothness selection and moderate number of
basis functions --- here we set \texttt{k\ =\ 15} for illustration.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{braya_low_k <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{15}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya, }\DataTypeTok{method =} \StringTok{"REML"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Before proceeding, we should perform a check to determine if the number
of basis functions requested is sufficient to capture the wiggliness in
the data. This test is provided by \texttt{gam.check()}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{gam.check}\NormalTok{(braya_low_k)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
#> 
#> Method: REML   Optimizer: outer newton
#> full convergence after 6 iterations.
#> Gradient range [-4.468978e-08,8.696688e-10]
#> (score -187.2524 & scale 0.000689802).
#> Hessian positive definite, eigenvalue range [0.5525272,43.5153].
#> Model rank =  15 / 15 
#> 
#> Basis dimension (k) checking results. Low p-value (k-index<1) may
#> indicate that k is too low, especially if edf is close to k'.
#> 
#>            k'   edf k-index p-value    
#> s(Year) 14.00  2.62    0.56  <2e-16 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

In the output, the column labelled \texttt{k\textquotesingle{}} is the
size of the basis used (the number of basis functions) --- note that it
is 14 and not the requested 15 functions as the constant basis function
has been removed as it is confounded with the model constant term, the
intercept. The column labelled \texttt{edf} indicates the
\emph{effective degrees of freedom} for the estimated smooth. The
\texttt{k-index} column contains the value of a test statistic for a
test of sufficient number of basis functions --- ideally, we'd like the
value of \texttt{k-index} to be close to or greater than 1. The quoted
\emph{p} value measures the support for the null hypothesis that enough
basis functions were available. In practial terms, the low
\texttt{k-index} and very low \emph{p} value suggest that the requested
number of basis functions used to fit the model was too low.

The solution is to increase \texttt{k}, say by doubling the original
value, and refit the model and perform the basis dimension check using
\texttt{gam.check()}, repeating this process until \texttt{k-index} is
close to or greater than 1 and the \emph{p} value is larger than a
desired threshold. In my experience, it may not be possible with some
data sets to increase \texttt{k} large enough to result in a
\texttt{k-index} \textgreater{} 1 and a \emph{p} value \textgreater{}
0.05, given the available. It is important to remember that the test
provided by \texttt{gam.check()} is only an heuristic one and shouldn't
be seen as infallible. If you find yourself increasing \texttt{k} to
large values relative to the number of samples in your data set without
achieving a \texttt{k-index} \textgreater{} 1 or \emph{p} \textgreater{}
0.05, then note the value in the \texttt{edf} column as you increase
\texttt{k}. If the \texttt{edf} of the model changes little as you
continue to increase \texttt{k}, this may be evidence that the test is
failing in this particular instance and that the estimated smooth is not
dependent on the value of \texttt{k} chosen.

\subsection{Accounting for heteroscedasticity due to time
averaging}\label{accounting-for-heteroscedasticity-due-to-time-averaging}

To proceed with the Braya-Sø example, we need to increase the basis
dimension (\texttt{k\ =\ 45}), fit using \texttt{method\ =\ "REML"}. As
discussed in the manuscript, we should also account for the non-constant
variance, or \emph{heteroscedasticity}, that may arise due to variation
in the amount of time (or ``lake years'') that each sample represents.
All else equal, we expect that samples that average more time have
smaller variance than those that average a smaller amount of time. To
include information about the expected heteroscedasticity in the GAM we
can use observational weights\footnote{This simple solution is for a
  Gaussian GAM. For models with other distributions, the exact
  interpretation of observational weights will vary. For such data,
  where the conditional mean and variance of the observations are
  linked, the use of location-scale, or distributional, models may be
  more apporpriate and informative.}. Here I use the
\texttt{sampleInterval} variable that I created earlier as the measure
of lake years per sample, and to avoid changing the model likelihood,
the weights use are the values of \texttt{sampleInterval} divided by the
mean of \texttt{sampleInterval}:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{braya_reml <-}\StringTok{ }\KeywordTok{gam}\NormalTok{(UK37 ~}\StringTok{ }\KeywordTok{s}\NormalTok{(Year, }\DataTypeTok{k =} \DecValTok{45}\NormalTok{), }\DataTypeTok{data =} \NormalTok{braya,}
                  \DataTypeTok{method =} \StringTok{"REML"}\NormalTok{,}
                  \DataTypeTok{weights =} \NormalTok{sampleInterval /}\StringTok{ }\KeywordTok{mean}\NormalTok{(sampleInterval))}
\KeywordTok{summary}\NormalTok{(braya_reml)}
\end{Highlighting}
\end{Shaded}
